Taking uncertainty in word embedding bias estimation seriously: a  bayesian

### 2023-06-01 paper/paperDraft.Rmd
literature remains

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '../')

library(ggplot2)
library(ggthemes)
library(gridExtra)
library(dplyr)
library(dagitty)
library(tidyverse)
library(magrittr)
library(kableExtra)
library(ggpubr)
library(ggExtra)
library(ggthemes)
library(ggforce)
library(rethinking)
library(latex2exp)
library(Hmisc)
library(grid)

@@

Conceptually, the idea is that if a particular harmful stereotype is learned in a particular embedding, then certain groups of words will be systematically closer to (or further from) each other. This gives rise to the idea of protected groups---for example, in guiding online search completion or recommendation, female words might require protection in that they should not be systematically closer to stereotypically female job names, such as "nurse", "librarian", "waitress", and male words require protection in that they should not be systematically closer to toxic masculinity stereotypes, such as "tough", "never complaining" or "macho".^[However, for some research-related purposes, such as the study of stereotypes across history [@Garg2017hundredYears], embeddings which do not protect certain classes may also be useful.]

@@

One of the first measures of bias has been developed in [@Bolukbasi2016man].  The general idea is that a certain topic is associated with a vector of real numbers (the topic "direction"), and the bias of a word is investigated by considering the projection of its corresponding vector on this direction. For instance, in [@Bolukbasi2016man],   the gender direction $\mathsf{gd}$   is obtained by taking the differences of the vectors corresponding to ten different gendered pairs (such as $\overrightarrow{she} - \overrightarrow{he}$ or $\overrightarrow{girl} - \overrightarrow{boy}$), and then identifying their principal component.^[Roughly, the principal component  is the vector obtained by projecting the data points on their  linear combination in a way that maximizes the variance of the projections.]  The gender bias of a word $w$ is then understood as $w$'s projection on the gender direction: $\vec{w} \cdot gd$ (which, after normalizing by dividing by $\lVert  w \rVert \,\lVert \mathsf{gd} \rVert$, is the same as cosine similarity).  Given a list  $N$ of  supposedly gender neutral words,^[We follow the methodology used in the debate in assuming that there is a class of words  identified as more or less neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose average similarity to the gender direction (or other protected words) is around 0. We will have more to say about this assumption when we describe our dataset construction.]  and the gender direction $\mathsf{gd}$,    the direct gender bias is defined as the average cosine similarity of the words in $N$ from $\mathsf{gd}$ ($c$ is a parameter determining how strict we want to be):
\begin{align*}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},\mathsf{gd})\vert^c}{\vert N \vert }
\end{align*}
\normalsize

@@

A measure of bias in word embeddings which does not employ similarities to  pre-calculated protected-class-level directions, the Word Embedding Association Test (\textsf{WEAT}), has been proposed in  [@Caliskan2017semanticsBiases]. The idea here is that the   bias between two sets of target words, $X$ and $Y$ (we call them protected words), should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we will call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related, and $B$ stereotypically female-related career words. The association difference for a particular word $t$ (belonging to either $X$ or $Y$) is:

@@

\noindent The authors use it as a test statistic in some tests, and the assumption that $X$ and $Y$ are of the same size is necessary for the statistic to make sense. In the final measure (of effect size), \textsf{WEAT}, means are taken, which makes it insensitive to size differences between $X$ and $Y$:
\begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

 \textsf{WEAT} is inspired by  the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology, and it uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans. In  [@Caliskan2017semanticsBiases] the authors argue that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. In [@Lauscher2019multidimensional] the methodology is  extended to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of cosine similarity does not make much difference, while the bias effects vary greatly across embedding models (interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia).
 A similar methodology is employed in [@Garg2018years]. The authors employ word embeddings trained on corpora from different decades to study the shifts in various biases.^[Strictly speaking, these authors use Euclidean distances and their differences, but the way they take averages and averages thereof is analogous, and so what we will have to say about pre-averaging leading to false confidence applies to this methodology as well.]

@@

\textsf{WEAT}, however, has been developed to investigate biases corresponding to a pair of supposedly opposing stereotypes, and so the question arises as to how to generalize the measure to contexts in which biases with respect to more than two stereotypical groups are to be measured. Such a generalization can be found in [@Manzini2019blackToCriminal]. The authors introduce Mean Average Cosine distance (\textsf{MAC}) as a measure of bias. Let $T = \{t_1, \dots, t_k\}$ be a set of protected words, and let each $A_j\in \mathcal{A}$ be a set of attributes stereotypically associated with a protected word where $\mathcal{A}$.
For instance, when biases related to religion are to be investigated, they use a dataset of the  format illustrated in Table \ref{tab:religionOriginal}. The measure is defined as follows:

@@

\begin{align*}
\mathsf{s}(t, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cosineDistance}(t,a) \\
\mathsf{mac}(T,\mathcal{A}) & = \frac{1}{\vert T \vert \,\vert \mathcal{A}\vert}\sum_{t \in T }\sum_{A_j \in \mathcal{A}}    \mathsf{s}(t,A_j)
\end{align*}

\noindent That is, for each  protected word $t\in T$, and each attribute set $A_j$, they first take the mean of distances for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words and all the protected classes.^[The authors'  code are in their github repository, [https://github.com/TManzini/DebiasMulticlassWordEmbedding](https://github.com/TManzini/DebiasMulticlassWordEmbedding).]

@@

ggplot(mac, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
                                                                 nudge_y = 0.02, size = 9 )+
  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.1),  linetype =2, arrow=arrow(), size=.03, color="blue")+geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.2), linetype =4, arrow=arrow(), size=.0, color="blue") + geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.3), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.1), linetype =3, arrow=arrow(), size=.05, color="chartreuse4")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.2), linetype =2, arrow=arrow(), size=.05, color="chartreuse4")+ annotate("text", x = c(.1,.4,.7,1.2), y = c(.45,.45,.45,.45),  label = c("T", "A[1]","A[2]", "A[3]") ,parse = TRUE, color=c("black", "blue","orangered","chartreuse4"),            size=10,  fontface="bold")+xlim(0,1.3)+theme_void(base_size = 25)

@@

Here is an example of results of debiasing that can be found in [@Manzini2019blackToCriminal] (Table \ref{tab:religionOriginal2}). It presents the results of applying the procedure with respect to the Religion dataset. \textsf{MAC} scores are given for the original datasets, and two resulting from two debiasing methods applied to the Reddit embedding. The scores for debiasing come with p-values for the comparison to the original \textsf{MAC}.

@@

The first conceptual question we should ask is whether   the initial \textsf{MAC} values lower than 1 indeed are  indicative of the presence of bias? Of course, thinking abstractly, 1 is the ideal distance for unrelated words. But clearly, there is variation in distances, which might lead to non-biased lists also having \textsf{MAC} smaller than 1. How much smaller?  What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high (i.e. close to 1) even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not large enough to consider it as bias. The authors, though, still aim to mitigate such "bias" to make the distance even larger. The question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.

The problem is that the original paper did not employ any control group of neutral attributes for comparison to obtain a more realistic gauge on how to understand \textsf{MAC} values. Later on, in our approach, we introduce such control word lists. One of them is a list of words we intuitively considered neutral (see the Appendix for word lists).

Moreover, it might be the case that words that have to do with human activities in general, even if unbiased, are systematically closer to the protected words than merely neutral words. This, again, casts doubt on whether comparing \textsf{MAC} to the abstractly ideal value of 1 is a methodologically sound idea. For this reason we also use a second list with intuitively non-stereotypical human attributes (again, see the Appendix for a word list).

@@

In Figures  (\ref{fig:empirical0}-\ref{fig:empirical3}) we look at the empirical distributions, while paying attention to such divisions. The horizontal lines represent the values of 1-\textsf{MAC} (that is, we now talk in terms of cosine similarity rather than cosine distance; this is a trivial linear transformation) that the authors considered indicative of bias for stereotypes corresponding to  given  word lists. For instance, in religion, \textsf{MAC} was .859 and considered a sign of bias, so we plot $0\pm (1-.859)\approx .14$ lines around similarity = 0 (that is, distance = 1). Notice that most distributions are quite wide, and the proportions of even neutral or human neutral words with similarities higher than the value of 1-\textsf{MAC} deserving debiasing according to the authors is quite high.

@@

To avoid such problems, later on, we employ control groups and in line with Bayesian methodology, use posterior distributions and  highest posterior density intervals instead of chasing single-point metrics based on pre-averaged data. But before we do so, let us spend some time explaining why pre-averaging is a sub-optimal strategy.

@@

The approaches we have been describing use  means of mean average cosine similarities to measure  similarity between protected words and attributes coming from harmful stereotypes. However, a closer inspection look at the individual values that are taken for the calculations, it turns out that there are quite a few  outliers and surprisingly dissimilar words. This problem becomes transparent when we examine the visualizations of the individual cosine distances, following the idea that one of the first steps in understanding data is to look at it. Let's start with inspecting two examples in Figures \ref{fig:muslim} and \ref{fig:priest}.

\begin{figure}[H]
```{r visMuslimd,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
religion <- read.csv("datasets/cleanedDatasets/religionRedditCleaned.csv")
muslimWords <- c("imam","islam","mosque","muslim","quran")
muslim <- religion %>% filter(protectedWord %in% muslimWords)
muslimClass <- muslim %>% filter(protectedWord == "muslim")
neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
muslimVis <- rbind(muslimVis,neutralSample,humanSample)

@@

visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 11)
```
\caption{Actual distances for the protected word "muslim". The most similar attributes tend to be the ones associated with it stereotypically, but then words associated with other stereotypes come closer than neutral or human predicates. }
\label{fig:muslim}
\end{figure}

\begin{figure}[H]
```{r visPriestd,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
religion <- read.csv("datasets/cleanedDatasets/religionRedditCleaned.csv")
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

@@

\caption{Actual distances for the protected word "priest", the situation is even less as expected: the nearest attributes are human attributes, and all there seems to be no clear pattern to the distances to other attributes.}
\label{fig:priest}
\end{figure}

With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results. We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.

\noindent
$s(X,Y,A,B)$ is the statistic used in the significance test, and the $p$-value is obtained by bootstrapping: it is the frequency of $s(X_i,Y_i,A,B)>s(X,Y,A,B)$ for all equally sized partitions $X_i, Y_i$ of $X\cup Y$. The effect size is computed by normalizing the difference in means as follows:

\footnotesize

\begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

@@

The t-tests they employ are run on average cosines used to calculate \textsf{MAC}.

The effect sizes reported by [@Caliskan2017semanticsBiases] for lists of words for which the embeddings are supposedly biased range from 2.06 to 1.81, and the reported $p$-values are in the range of $10^{-7}-10^{-2}$ with one exception for Math vs arts, where it is $.018$. The question is, are those results meaningful?

One way to answer this question is to think in terms of null models. If the words actually are samples from two populations with equal mean, how often would we see effect sizes in this range? How often would we reach the $p$-values that the authors reported?

We start our exploration with considering a list of 16 protected words, each class containing 8, with 16 attributes, 8 in each attribute class. In each particular sample we will investigate the consequence of the model being in fact null: we draw the cosine distances, every time, from the $\mathsf{Normal}(0,.08)$ distribution. $0.08$ is approximately the empirical standard deviation observed in fairly large samples of neutral words, and 16 is the sample size used in the WEAT7 word list, which is not much different from the other sample sizes in word lists used by [@Caliskan2017semanticsBiases]

@@

plotStat <- ggplot(combinations) + geom_density(aes(x = statistic))+
geom_vline(xintercept = ourStatistic, linewidth = .5, 
           lty = 2, alpha = .5)+theme_tufte(base_size = 10)+
  theme(plot.title.position = "plot", axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) +
  labs(title = "Boostrapped distribution of test statistics",
       subtitle = "Random null model with N(0,.008) and n = 16")+
   annotate("label", x = ourStatistic, y = 2,
            label =  paste(round(ourStatistic,2),
                           ", p=", 0.011, sep = ""), size = 3)+
  xlim(-2,2)+xlab("test statistic")+ylab("density")

plotEff <- ggplot(combinations) + geom_density(aes(x = es))+
  xlab("effect size")+ylab("density")+
  geom_vline(xintercept = ourEffectSize, linewidth = .5, 
             lty = 2, alpha = .5)+theme_tufte(base_size = 10)+
  theme(plot.title.position = "plot", axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) +
  labs(title = "Boostrapped distribution of effect sizes",
       subtitle = "Random null model with N(0,.008) and n = 16")+
  annotate("label", x = ourEffectSize, y = .5,
           label = paste(round(ourEffectSize,2),
                         ", p=", 0.0086, sep = "" ), 
           size = 3)+xlim(-2,2)

plotCaliskanDist <- ggplot(tDF) + 
  geom_point(aes(x = Avalue, y = Bvalue, color = target, shape = target
                 ), size = 3, alpha = .8)+
  theme_tufte(base_size = 10)+
  ggtitle("Cosine distances by word group in  our null dataset")+
  theme(plot.title.position = "plot", 
        legend.position = c(0.08, 0.89),
        legend.background = element_rect(fill="lightblue", 
         size=0.01, linetype="solid"))

@@

So let's draw one sample of this type. We calculate the s-values, the test statistics and the effect sizes. Then, following the original methodology, we obtain bootstrapped distributions of the effect sizes and the test statistics by calculating the same for each possible equal split of the initial random data set. The observed test statistic is `r ourStatistic` and `r ourEffectSize`. The bootstrapped distributions of the test statistics and effect sizes are illustrated in Figure \ref{fig:caliskanCalc}. Quite notably both (two-sided) $p$ values are uneven and rather low.

\begin{figure}[H]
```{r fig:caliskanCalcB2B,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
grid.arrange(plotStat,plotEff)
```

\caption{Bootstrapped distributions of test statistics and effect sizes in a random sample given the null hypothesis.}
\label{fig:caliskanCalc}
\end{figure}

>>>>

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '../')
library("viridis") 
library(ggplot2)
library(ggthemes)
library(gridExtra)
library(dplyr)
library(dagitty)
library(tidyverse)
library(magrittr)
library(kableExtra)
library(ggpubr)
library(ggExtra)
library(ggthemes)
library(ggforce)
library(rethinking)
library(latex2exp)
library(Hmisc)
library(grid)

@@

The underlying intuition is that if a particular harmful stereotype is learned in a particular embedding, then certain groups of words will be systematically closer to (or further from) each other. This gives rise to the idea of protected groups---for example, in guiding online search completion or recommendation, female words might require protection in that they should not be systematically closer to stereotypically female job names, such as "nurse", "librarian", "waitress", and male words require protection in that they should not be systematically closer to toxic masculinity stereotypes, such as "tough", "never complaining" or "macho".^[However, for some research-related purposes, such as the study of stereotypes across history [@Garg2017hundredYears], embeddings which do not protect certain classes may also be useful.]

@@

One of the first measures of bias has been developed in [@Bolukbasi2016man].  The general idea is that a certain topic is associated with a vector of real numbers (the topic "direction"), and the bias of a word is investigated by considering the projection of its corresponding vector on this direction. For instance, in [@Bolukbasi2016man],   the gender direction $\mathsf{gd}$   is obtained by taking the differences of the vectors corresponding to ten different gendered pairs (such as $\overrightarrow{she} - \overrightarrow{he}$ or $\overrightarrow{girl} - \overrightarrow{boy}$), and then identifying their principal component.^[Roughly, the principal component  is the vector obtained by projecting the data points on their  linear combination in a way that maximizes the variance of the projections.]  The gender bias of a word $w$ is then understood as $w$'s projection on the gender direction: $\vec{w} \cdot gd$ (which, after normalizing by dividing by $\lVert  w \rVert \,\lVert \mathsf{gd} \rVert$, is the same as cosine similarity).  Given a list  $N$ of  supposedly gender neutral words,^[We follow the methodology used in the debate in assuming that there is a class of words  identified as more or less neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose average similarity to the gender direction (or other protected words) is around 0. See our list in Appendix \ref{app:custom} and a brief discussion in Subsection \ref{subsec:interpretability}.]  and the gender direction $\mathsf{gd}$,    the direct gender bias is defined as the average cosine similarity of the words in $N$ from $\mathsf{gd}$ ($c$ is a parameter determining how strict we want to be):
\begin{align*}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},\mathsf{gd})\vert^c}{\vert N \vert }
\end{align*}
\normalsize

@@

A measure of bias in word embeddings which does not proceed by identifying bias directions (such as a gender vector), the Word Embedding Association Test (\textsf{WEAT}), has been proposed in  [@Caliskan2017semanticsBiases]. The idea here is that the   bias between two sets of target words, $X$ and $Y$ (we call them protected words), should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we will call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related, and $B$ stereotypically female-related career words. The association difference for a particular word $t$ (belonging to either $X$ or $Y$) is:

@@

\noindent The intention is that large values of $s$ scores suggest systematic differences between how $X$ and $Y$ are related to $A$ and $B$, and therefore are indicative of the presence of bias. The authors use it as a test statistic in some tests,^[Note their method assumes $X$ and $Y$ are of the same size.] and the final measure of effect size, \textsf{WEAT}, is constructed by taking means of these values and standardizing:
\begin{align} \label{eq:weat}
\mathsf{WEAT}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

 \textsf{WEAT} is inspired by  the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology, and in some applications it uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans. In  [@Caliskan2017semanticsBiases] the authors argue that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. In [@Lauscher2019multidimensional] the methodology is  extended to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of cosine similarity does not make much difference, while the bias effects vary greatly across embedding models.^[Interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia.]
 A similar methodology is employed in [@Garg2018years]. The authors employ word embeddings trained on corpora from different decades to study the shifts in various biases through the century.^[Strictly speaking, these authors use Euclidean distances and their differences, but the way they take averages and averages thereof is analogous, and so what we will have to say about pre-averaging leading to false confidence applies to this methodology as well.]

@@

\textsf{WEAT} has been developed to investigate biases corresponding to a pair of supposedly opposing stereotypes,  so the question arises as to how to generalize the measure to contexts in which biases with respect to more than two stereotypical groups are to be measured. Such a generalization can be found in [@Manzini2019blackToCriminal]. The authors introduce Mean Average Cosine distance (\textsf{MAC}) as a measure of bias. Let $T = \{t_1, \dots, t_k\}$ be a set of protected words, and let each $A_j\in \mathcal{A}$ be a set of attributes stereotypically associated with a protected word where $\mathcal{A}$.
For instance, when biases related to religion are to be investigated, they use a dataset of the  format illustrated in Table \ref{tab:religionOriginal}. The measure is defined as follows:

@@

\begin{align*}
\mathsf{s}(t, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cosineDistance}(t,a) \\
\mathsf{MAC}(T,\mathcal{A}) & = \frac{1}{\vert T \vert \,\vert \mathcal{A}\vert}\sum_{t \in T }\sum_{A_j \in \mathcal{A}}    \mathsf{s}(t,A_j)
\end{align*}

\noindent That is, for each  protected word $t\in T$, and each attribute set $A_j$, they first take the mean of distances for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words and all the protected classes.^[The authors'  code is available through their github repository at  [https://github.com/TManzini/DebiasMulticlassWordEmbedding](https://github.com/TManzini/DebiasMulticlassWordEmbedding).]

@@

ggplot(mac, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,                               nudge_y = 0.02, size = 9 )+
  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.1),  linetype =2, arrow=arrow(), size=.03, color="blue")+
  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.2),
               linetype =4, arrow=arrow(), size=.0, color="blue") + geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.3), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+ 
  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.1), linetype =3, arrow=arrow(), size=.05, color="chartreuse4")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.2), linetype =2, arrow=arrow(), size=.05, color="chartreuse4")+ 
  annotate("text", x = c(.1,.4,.7,1.1), y = c(.45,.45,.45,.45),  label = c("T", "A[1]","A[2]", "A[3]") ,parse = TRUE, color=c("black", "blue","orangered","chartreuse4"),            size=10,  fontface="bold")+xlim(0,1.3)+theme_void(base_size = 25)

@@

Table \ref{tab:religionOriginal2} contains  an example of \textsf{MAC} scores (and $p$ values, we explain how these are obtained in Subsection \ref{subsec:problems}) before and after  deploying two debiasing methods to the Reddit embedding, where the score is calculated using the Religion word lists from  [@Manzini2019blackToCriminal]. For our purpose the details of the debiasing method are not important: what matters is that \textsf{MAC} is used in the evaluation of these methods.

@@

The first question we should ask is whether   the initial \textsf{MAC} values lower than 1 indeed are  indicative of the presence of bias? Thinking abstractly, 1 is the ideal distance for unrelated words. But in fact there is some variation in distances, which might lead to non-biased lists also having \textsf{MAC} scores smaller than 1. How much smaller?  What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high (i.e. close to 1) even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not large enough to claim the presence of a bias to start with. The authors, though, still aim to mitigate it by making the distances involved in the \textsf{MAC} calculations even larger. The question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.

The problem is that the original paper did not employ any control group of neutral attributes for comparison to obtain a more realistic gauge on how to understand \textsf{MAC} values. Later on, in our approach, we introduce such control word lists. One of them is a list of words we intuitively considered neutral.  Moreover, it might be the case that words that have to do with human activities in general, even if unbiased, are systematically closer to the protected words than merely neutral words. This, again, casts doubt on whether comparing \textsf{MAC} to the abstractly ideal value of 1 is a methodologically sound idea. For this reason we also use a second list with intuitively non-stereotypical human attributes.^[See  Appendix \ref{app:custom} for the word lists.]

@@

In Figures  (\ref{fig:empirical0}-\ref{fig:empirical3}) we look at the empirical distributions, while paying attention to such divisions. The horizontal lines represent the values of $1 - \mathsf{MAC}$ (that is, we now talk in terms of cosine similarity rather than cosine distance) that the authors considered indicative of bias for stereotypes corresponding to  given  word lists. For instance, in religion, \textsf{MAC} was .859, which  was considered a sign of bias, so we plot $0\pm (1-.859)\approx .14$ lines around similarity = 0 (that is, distance = 1). Notice that most distributions are quite wide, and the proportions of even neutral or human neutral words with similarities higher than the value of $1 - \mathsf{MAC}$ deserving debiasing according to the authors are quite high.

@@

To avoid the problem brought up in this subsection, we employ control groups and in line with Bayesian methodology, use posterior distributions and  highest posterior density intervals instead of chasing single-point metrics based on pre-averaged data. Before we do so, we first explain why pre-averaging and chasing single-number metrics is a sub-optimal strategy.

@@

The approaches we have been describing use  means of mean average cosine similarities to measure  similarity between protected words and attributes coming from harmful stereotypes. But once we take a look at the individual values, it turns out that the raw data variance is rather high, and there are quite a few  outliers and surprisingly dissimilar words. This problem becomes transparent when we examine the visualizations of the individual cosine distances, following the idea that one of the first steps in understanding data is to look at it. Let's start with inspecting two examples of such visualizations in Figures \ref{fig:muslim} and \ref{fig:priest} (we also include neutral and human predicates to make our point more transparent). Again, we emphasize that **we do not condone the associations which we are about to illustrate.**

\begin{figure}[H]
```{r visMuslimd,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%",message =FALSE, warning = FALSE}
religion <- read.csv("datasets/cleanedDatasets/religionRedditCleaned.csv")
muslimWords <- c("imam","islam","mosque","muslim","quran")
muslim <- religion %>% filter(protectedWord %in% muslimWords)
muslimClass <- muslim %>% filter(protectedWord == "muslim")
neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
muslimVis <- rbind(muslimVis,neutralSample,humanSample)

@@

visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 11)
```
\caption{Actual distances for the protected word \textsf{muslim}. }
\label{fig:muslim}
\end{figure}

\begin{figure}[H]
```{r visPriestd,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%", warning = FALSE, message = FALSE}
religion <- read.csv("datasets/cleanedDatasets/religionRedditCleaned.csv")
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

@@

\caption{Actual distances for the protected word \textsf{priest}.}
\label{fig:priest}
\end{figure}

As transparent in Figures \ref{fig:muslim} and \ref{fig:priest}, for the protected word \textsf{muslim}, the most similar attributes tend to be the ones associated with it stereotypically, but then words associated with other stereotypes come closer than neutral or human predicates. For the protected word \textsf{priest}, the situation is even less as expected: the nearest attributes are human attributes, and all there seems to be no clear pattern to the distances to other attributes.

The general phenomenon that makes us skeptical about running statistical tests on pre-averaged data is that raw datasets of different variance can result in the same pre-averaged data and consequently the same single-number metric. In other words, a method that proceeds this way is not very sensitive to the real sample variance.

Let us illustrate how this problem arises in the context of \textsf{WEAT}. Once a particular  $s(X,Y,A,B)$ is calculated, the question arises as to whether a value that high could have arisen by chance. To address the question, each $s(X,Y,A,B)$ is used in the original paper to generate a  $p$-value by bootstrapping. The $p$-value  is the frequency of how often it is the case that  $s(X_i,Y_i,A,B)>s(X,Y,A,B)$ for sampled equally sized partitions $X_i, Y_i$ of $X\cup Y$.  The WEAT score is then computed by standardizing the difference in means of means by dividing by the standard deviation of means, see equation \eqref{eq:weat}.

<!-- \vspace{-2mm} -->

<!-- \footnotesize  -->

<!-- \begin{align} -->
<!-- \mathsf{bias}(A,B) & = \frac{ -->
<!-- \mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y})  -->
<!-- }{ -->
<!-- \sigma(\{s(w,A,B)\}_{w\in X\cup Y}) -->
<!-- } -->
<!-- \end{align} -->

@@

\noindent The \textsf{WEAT} scores reported by [@Caliskan2017semanticsBiases] for lists of words for which the embeddings are supposedly biased range from 2.06 to 1.81, and the reported $p$-values are in the range of $10^{-7}-10^{-2}$ with one exception for *Math vs Arts*, where it is $.018$.

The question is, are those results meaningful? One way to answer this question is to think in terms of null generative models. If the words actually are samples from two populations with equal means, how often would we see \textsf{WEAT} scores in this range? How often would we reach the $p$-values that the authors reported?

Imagine there are two groups of protected words, each of size 8, and two groups of stereotypical attributes, of the same size.^[16 is the sample size used in the WEAT7 word list, which is not much different from the other sample sizes in word lists used by [@Caliskan2017semanticsBiases]).] Each such a collection of samples, as far as our question is involved, is equivalent to a sample of $16^2$ cosine distances. Further, imagine there really is no difference between these groups of word and the  model is in fact null. That is,  we draw the cosine distances from the $\mathsf{Normal}(0,.08)$ distribution.^[$.08$ is approximately the empirical standard deviation observed in fairly large samples of neutral words.]

@@

plotStat <- ggplot(combinations) + geom_density(aes(x = statistic))+
geom_vline(xintercept = ourStatistic, linewidth = .5, 
           lty = 2, alpha = .5)+theme_tufte(base_size = 10)+
  theme(plot.title.position = "plot", axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) +
  labs(title = "Boostrapped distribution of test statistics: s values")+
   annotate("label", x = ourStatistic, y = 2,
            label =  paste(round(ourStatistic,2),
                           ", p=", 0.011, sep = ""), size = 3)+
  xlim(-2,2)+xlab("s values")+ylab("density")

plotEff <- ggplot(combinations) + geom_density(aes(x = es))+
  xlab("WEAT")+ylab("density")+
  geom_vline(xintercept = ourEffectSize, linewidth = .5, 
             lty = 2, alpha = .5)+theme_tufte(base_size = 10)+
  theme(plot.title.position = "plot", axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) +
  labs(title = "Boostrapped distribution of effect sizes: WEAT scores")+
  annotate("label", x = ourEffectSize, y = .5,
           label = paste(round(ourEffectSize,2),
                         ", p=", 0.0086, sep = "" ), 
           size = 3)+xlim(-2,2)

plotCaliskanDist <- ggplot(tDF) + 
  geom_point(aes(x = Avalue, y = Bvalue, color = target, shape = target
                 ), size = 3, alpha = .8)+
  theme_tufte(base_size = 10)+
  ggtitle("Cosine distances by word group in  our null dataset")+
  theme(plot.title.position = "plot", 
        legend.position = c(0.08, 0.89),
        legend.background = element_rect(fill="gray89", 
         size=0.01, linetype="solid"))+
  scale_colour_manual(values = c("darkorange3", "grey40"))+
  xlab("distance to A")+ylab("distance to B")

@@

In Figure \ref{fig:caliskanCalc} we illustrate one iteration of the procedure.  We draw one such sample of size $16^2$. Then we actually list all possible ways to split the 16 words in two equal sets (each such a split is one bootstrapped sample) and for each of them we calculate the \textsf{s} values and \textsf{WEAT}. What are the resulting distributions of  $\textsf{s}$ scores and what $p$-values do they lead to?
What are the resulting effect sizes for each bootstrapped sample, and how often can we get an effect size as large as the ones reported in the original paper?

In the bootstrapped samples we would rather expect low \textsf{s} values and low \textsf{WEAT}: after all, these are just random permutations of random distances all of which are drawn from the same null distribution. Let's take a look at one such a bootstrapped sample.
